\documentclass{llncs}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc}


%\usepackage{url}
\usepackage{multirow}
\usepackage{listings}
\usepackage{amsmath}  % for equation*
\usepackage{array}    % for tabular
\usepackage{verbatim} % for comment
\usepackage{wrapfig}
\usepackage[final]{microtype}
\usepackage[pdfborder={0 0 0}]{hyperref}
\usepackage{tabularx}

\newcommand\forAll[1]{\forall \, #1 \, . \,}
\newcommand\forAllII[2]{\forall \, #1 \, #2 \, . \,}

\newcommand\propno[1]{(\emph{#1})}
\newcommand\hs[1]{\texttt{#1}}

% \raggedbottom

\lstnewenvironment{code}[1][]
  {\noindent
   \vspace{-0.5\baselineskip}
   \lstset{basicstyle=\ttfamily,
           frame=single,
           language=Haskell,
           keywordstyle=\color{black},
           #1}
   \fontsize{8pt}{8pt}\selectfont}
  {}

\newcommand\NOTE[1]{} % \mbox{}\marginpar{\fontsize{8pt}{8pt}\selectfont\raggedright\hspace{0pt}\emph{#1}}}

\def\maindocument{} % To tell tikz images that they are not stand alone

\begin{document}

\title{TIP: Tons of Inductive Problems}

\author{Koen Claessen \and Moa Johansson \and Dan Ros\'en \and Nicholas Smallbone}
\institute{Department of Computer Science and Engineering, Chalmers University of Technology
	\email{\{koen,jomoa,danr,nicsma\}@chalmers.se}
	}

\authorrunning{Claessen, Johansson, Ros\'en, Smallbone}
\titlerunning{}

\maketitle

\begin{abstract}
This paper describes initial efforts to collect benchmarks for 
inductive theorem provers which can be shared among developers 
and users. The recent increase of interest in automated 
inductive theorem proving (in special-purpose provers such 
as IsaPlanner, Zeno and HipSpec, in SMT-solvers such as CVC4, 
and in auto-active provers like Dafny) has also increased the 
demands for evaluation and comparison between these systems. Our 
benchmark suite aims to become the standard shared benchmark 
suite for inductive theorem provers. We expect the benchmark 
suite to continuously grow as more problems are submitted by the 
community. New challenge problems will promote new developments 
of provers which will greatly benefit both developers and users 
of inductive theorem provers.
\end{abstract}

\section{Introduction}

We have recently seen increased interest in inductive theorem proving, both with specialised provers such as IsaPlanner, Zeno and HipSpec \cite{hipspecCADE,dixon2007isaplanner,zeno}, SMT-solvers such as CVC4 \cite{cvc4}, the auto-active prover Dafny \cite{dafny}, recent work on the first-order SPASS prover \cite{SPASSInduction}, as well as some support in proof assistants \cite{hipster,acl2}.

To ease evaluation and development, and compare the relative strengths of the different systems, it is important to have good standard benchmarks. The contribution of this paper is an accessible standard benchmark suite for inductive theorem provers which can be extended by users and developers. The benchmarks are publicly available at:
\begin{center}
\url{https://github.com/tip-org/benchmarks}
\end{center}
We have so far collected 205 problems in the TIP benchmark suite. We invite the community to submit additional problems and challenges and expect the collection to continuously grow and provide new challenges for developers.

%% We do not yet discuss any special-purpose language features for expressing inductive problems in general. The benchmarks are therefore currently expressed in a subset of the WhyML specification language from the Why3 program verification system \cite{boogie11why3,Why3}. We have chosen this format as the Why3 system already provides translation tools from WhyML to various other common formats, such as SMT-LIB and versions of TPTP.
%% In addition, the WhyML language is also easy for humans to read. The Why3 system relies on external provers to discharge proof obligations and is already connected to a wide variety of automated and interactive provers, including  Alt-Ergo, CVC3, Z3, E, SPASS, Vampire, Coq and PVS. We hope that the new benchmark suite will encourage the connection of automated inductive theorem provers to Why3.

\section{The Benchmark Format}
\label{sec:format}
The benchmarks are expressed in a variant of SMT-LIB \cite{BST10},
extended with support for algebraic datatypes and higher-order
functions. The format is described in detail in \cite{tiplang}.
Starting from the basic SMT-LIB format, we import the following
features from existing systems:
\begin{itemize}
\item Algebraic datatypes, which are declared using the
  \verb|declare-datatypes| syntax as supported in Z3 and CVC4.
\item Recursive function definitions, which use the
  \verb|define-funs-rec| syntax implemented in CVC4 and proposed for
  SMT-LIB 2.5 \cite{BST10}.
\item Polymorphic functions, using the proposed \verb|par| syntax
  \cite{cvc4parPR}, which is implemented in a version of CVC4.
\end{itemize}

We then add more features of our own, which are specific to TIP:
\begin{itemize}
\item In the standard \verb|declare-datatypes| syntax, functions over
  algebraic datatypes are defined using projection functions like
  \verb|head| and \verb|tail|. We add a pattern-matching syntax,
  which is more convenient for many provers.
\item Many of our benchmark problems use higher-order functions,
  such as \verb|map|. We add syntax for lambda functions and
  higher-order functions.
\item Many inductive provers treat the goal specially, as opposed to
  SMT-LIB which expresses the goal as a negated axiom. We add a
  construct \verb|(assert-not p)| which declares \verb|p| as the goal;
  it is semantically equivalent to \verb|(assert (not p))|.
\end{itemize}

We do not necessarily expect every prover to support TIP natively.
Instead, we are working on a tool which can translate TIP problems to
and from a variety of other formats. Currently our tool can translate
TIP problems to a CVC4-compatible version of SMT-LIB or to WhyML, and
can compile Haskell programs into TIP properties. It can also perform
a number of transformations for tools which do not support the full
TIP language, such as removing higher-order functions by
defunctionalisation \cite{defunc}.

\subsection*{Example}
As an example of what the benchmarks look like, we show \texttt{prop12} from the IsaPlanner benchmark set (see Section~\ref{sec:isap} below), which states that the functions \texttt{drop} and \texttt{map} distribute over one another: \begin{center} \texttt{drop n (map f xs) = map f (drop n xs)}.\end{center}
We declare two simple algebraic datatypes representing natural numbers and polymorphic lists.

\begin{code}
(declare-datatypes (a)
  ((list (nil) (cons (head a) (tail (list a))))))
(declare-datatypes () ((Nat (Z) (S (p Nat)))))
\end{code}

Next, we declare two recursive functions: \texttt{drop}, which recursively drops a given number of elements from the front of the list and \texttt{map}, which is a higher-order function applying the given unary function \texttt{f} to each element of a list.

\begin{code}
(define-funs-rec
  ((par (a b) (map ((f (=> a b)) (xs (list a))) (list b))))
  ((match xs
     (case nil (as nil (list b)))
     (case (cons y ys) (cons (@ f y) (map f ys))))))
(define-funs-rec
  ((par (a) (drop ((n Nat) (xs (list a))) (list a))))
  ((match n
     (case Z xs)
     (case (S m)
       (match xs
         (case nil xs)
         (case (cons y ys) (drop m ys)))))))
\end{code}

The definition illustrates several features of TIP:
\begin{itemize}
  \item Polymorphism: \verb|par| introduces type variables.
  \item Higher-order functions: \verb|(=> a b)| is the type of
    functions from \verb|a| to \verb|b|, while \verb|@| applies
    a first-class function to its argument.
  \item Pattern-matching.
\end{itemize}

Finally, the benchmark problem itself is declared with the keyword \texttt{assert-not}:

\begin{code}
(assert-not
  (par (a b)
    (forall ((n Nat) (f (=> a b)) (xs (list a)))
      (= (drop n (map f xs)) (map f (drop n xs))))))
(check-sat)
\end{code}
Each benchmark file is stand-alone and only contains one property.

\section{Sample Benchmarks}
In this section we give a short overview of  some the problems currently available in the repository. We expect more to be added.

\subsection{IsaPlanner's Rippling and Case-Analysis Benchmarks}
\label{sec:isap}
This set of 85 problems comes from the evaluation of IsaPlanner's rippling-heuristic for guiding rewriting in inductive proofs in the context of functions with case- and if-statements \cite{IsaPcase}. It has been used in the evaluation of many of the recent inductive theorem provers and includes theorems about lists, natural numbers and binary trees. The problems are relatively easy, most of the theorems can be proved by structural induction using only the function definitions and only 15 require auxiliary lemmas to be discovered.

\subsection{Productive Use of Failure Benchmarks}
This is another benchmark suite which has been used to evaluate several recent provers. It consists of 50 theorems about lists and natural numbers and originates from evaluation of techniques for discovering auxiliary lemmas in the CLAM prover \cite{productiveuse}. The original paper did not provide definitions for the functions used in the benchmarks, so the definitions provided here come from the evaluation of the HipSpec system \cite{hipspecCADE}. These proofs are generally a bit harder, and may require additional lemmas to be found and proved (by another induction) or generalisation of the conjecture in order to strengthen the inductive hypothesis.

\subsection{New TIP Benchmarks}
This set contains 70 new benchmarks including, amongst others,
properties of the Agda standard
library\footnote{\url{https://github.com/agda/agda-stdlib}}
implementation of integers on top of natural numbers, problems about natural numbers in binary representation, problems about regular expressions, binary search trees, grammars and skew heaps. The problems about regular expressions, grammars and heaps have to our knowledge not been fully automated yet and are offered as challenges!

\section{Contribute to TIP}
We invite the theorem proving community to contribute additional inductive
benchmarks and challenge problems to TIP. Instructions for how to submit
problems can be found in the README file for the repository
(\url{https://github.com/tip-org/benchmarks}.)

We are developing a toolchain for translating to and from our format. The
development is in its own repository
(\url{https://github.com/tip-org/tools}).
The tool can currently read in problems in TIP format or Haskell, and
output TIP, SMT-LIB and WhyML, with some caveats:
\begin{itemize}
  \item The generated SMT-LIB uses \verb|declare-datatypes|,
    \verb|define-funs-rec| and polymorphism.
  \item The generated WhyML makes no special effort to pass Why3's
    termination checker.
\end{itemize}
We encourage the community to request and contribute additional input
and output formats to our toolchain.

\section{Conclusion and Further Work}
TIP is intended to be a standard benchmark suite for developers and users of inductive theorem provers. We hope that this initiative will ease comparison and evaluation of systems and spur further collaboration and development by attracting submissions of additional challenge problems from the community.

In addition to serving as a standard benchmark suite for inductive provers, the TIP benchmarks may also be useful for developers of theory exploration systems. Theory exploration is a technique for automatically discovering interesting conjectures about a given set of functions and datatypes, and is used in for example the HipSpec prover to discover lemmas. The TIP benchmarks can be compared to the output from the theory explorer in precision/recall analysis to assess the quality and interestingness of the conjectures generated. A good theory exploration system may also be used to generate new benchmarks for TIP.

It is important to have good tool support if TIP is to be used by the
community. We plan to develop our tool further to support more input
and output formats, as well as various strategies for encoding
features of TIP for provers that do not support the full language.

%The TIP benchmarks are currently expressed in a subset of WhyML. If we want to use the Why3 translation tools to other formats, there are some restrictions. Functions are, for instance, required to be provably terminating by the Why3 termination checker, which might not always be the case (e.g., it fails to prove termination for many functions that are not structurally recursive).
%In the future, we therefore expect to support a richer language, and will provide tools to help developers use the TIP format.
In the future, we may want to extend the language to support a richer
variety of problems. For example,
we may want to include problems about lazy functions and will then need to extend the format to allow non-terminating functions.
%We may in the future perhaps also want to include problems featuring functions in a programming language with lazy evaluation.
Another extension is to also allow non-theorems among the benchmarks, for evaluation of tools for counter-example finding.  We see this paper as a starting point to encourage further discussion in the inductive theorem proving community about what future format(s) for inductive problems should look like.



\bibliographystyle{plain}
\bibliography{bibfile}

\end{document}
